---
title: "Class09 Interactive Session: Unsupervised Learning Analysis of Human Breast Cancer Cells"
author: "Madeline R. Luth"
date: "5/1/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background Information

The full session instructions can be accessed at: https://bioboot.github.io/bggn213_S19/class-material/lab-9-bggn213-WEBX.html#7_prediction

The goal of this hands-on session is to explore a complete analysis using the unsupervised learning techniques covered in the last class. Extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

## Exploratory Data Analysis

```{r}

# read the data in
wisc.df <- read.csv("https://bioboot.github.io/bggn213_W19/class-material/WisconsinCancer.csv")

```

There are 569 patients in the dataset (# of observations).
Can also be determined using nrow().

```{r}

head(wisc.df)

```

```{r}
table(wisc.df$diagnosis)
```

There are 212 malignant diagnoses and 357 benign.

Can also use filter feature in the R Studio global environment.

```{r}

length(grep("_mean", colnames(wisc.df)))

```

There are 10 features in the data that are suffixed with "_mean".

```{r}

# read in only columns 3-32 (there are NAs at the end of the file and we don't need ID and diagnosis string values in this analysis)
wisc.data <- as.matrix(wisc.df[,3:32])

```

```{r}

# Set the row names of wisc.data
row.names(wisc.data) <- wisc.df$id
head(wisc.data)

```

```{r}

# create diagnosis vector for later
diagnosis <- wisc.df$diagnosis
head(diagnosis)
  
```

## Principal Component Analysis

The next step in your analysis is to perform principal component analysis (PCA) on wisc.data.

It is important to check if the data need to be scaled before performing PCA. Recall two common reasons for scaling data include:

The input variables use different units of measurement.
The input variables have significantly different variances.

```{r}

# Check column means and standard deviations
colMeans(wisc.data)
apply(wisc.data,2,sd)

```

The data need to be scaled prior to generating PCA.

```{r}

# Perform PCA on wisc.data
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
summary(wisc.pr)

```

From your results, what proportion of the original variance is captured by the first principal components (PC1)?

0.4427

How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3

How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7

```{r}

biplot(wisc.pr)

```

This plot is generally uninformative. Will need to make a better one.

```{r}

#plot PC1 and PC2
plot(wisc.pr$x[,1], wisc.pr$x[,2], xlab="PC1", ylab="PC2", col = diagnosis)

```


```{r}

# Plot PC1 and PC3
plot(wisc.pr$x[,1], wisc.pr$x[,3], xlab="PC1", ylab="PC3", col = diagnosis)
```

Because principal component 2 explains more variance in the original data than principal component 3, you can see that the first plot has a cleaner cut separating the two subgroups.

Overall, the plots indicate that principal component 1 is capturing a separation of malignant from benign samples.

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components. Assign this to a variable called pve and create a plot of variance explained for each principal component.

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / length(wisc.pr)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}

# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

```

There are quite a few CRAN packages that are helpful for PCA. This includes the factoextra package. For example:

```{r}

## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)

```

```{r}

# Access the contribution of one variable on PC1 
wisc.pr$rotation["concave.points_mean",1]

# Sort the loadings values by their absolute contribution to PC1
sort( abs(wisc.pr$rotation[,1]) )

```

```{r}
wisc.pr$rotation["radius_mean",1]
wisc.pr$rotation["smoothness_se",1]
```


## Hierarchical Clustering

The goal of this section is to do hierarchical clustering of the observations. Recall from our last class that this type of clustering does not assume in advance the number of natural groups that exist in the data.

As part of the preparation for hierarchical clustering, the distance between all pairs of observations are computed. Furthermore, there are different ways to link clusters together, with single, complete, and average being the most common linkage methods.

```{r}

# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(d=data.dist, method ="complete")

```

```{r}

plot(wisc.hclust)
abline(h=19, col="red", lty=2)

```

In this section, you will compare the outputs from your hierarchical clustering model to the actual diagnoses. Normally when performing unsupervised learning like this, a target variable (i.e. known answer or labels) isn't available. We do have it with this dataset, however, so it can be used to check the performance of the clustering model.

When performing supervised learning - that is, when you're trying to predict some target variable of interest and that target variable is available in the original data - using clustering to create new features may or may not improve the performance of the final model.

This exercise will help you determine if, in this case, hierarchical clustering provides a promising new feature.

```{r}

# Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable wisc.hclust.clusters.
wisc.hclust.clusters <- cutree(wisc.hclust, h = 19)
# We can use the table() function to compare the cluster membership to the actual diagnoses.
table(wisc.hclust.clusters, diagnosis)

```

Here we picked four clusters and see that cluster 1 largely corresponds to malignant cells (with diagnosis values of 1) whilst cluster 3 largely corresponds to benign cells (with diagnosis values of 0).

## K-means Clustering

In this section, you will create a k-means clustering model on the Wisconsin breast cancer data and compare the results to the actual diagnoses and the results of your hierarchical clustering model. If you are running a little behind feel free to skip ahead to section 5 otherwise if you find you are flying through things please take some time to see how each clustering model performs in terms of separating the two diagnoses and how the clustering models compare to each other.

Create a k-means model on wisc.data, assigning the result to wisc.km. Be sure to create 2 clusters, corresponding to the actual number of diagnosis. Also, remember to scale the data (with the scale() function and repeat the algorithm 20 times (by setting setting the value of the nstart argument appropriately). Running multiple times such as this will help to find a well performing model.

```{r}
wisc.km <- kmeans(data.scaled, centers= 2, nstart= 20)
table(wisc.km$cluster, diagnosis)
```

## Combining Methods

In this final section, you will put together several steps you used earlier and, in doing so, you will experience some of the creativity and open endedness that is typical in unsupervised learning.

Recall from earlier sections that the PCA model required significantly fewer features to describe 70%, 80% and 95% of the variability of the data. In addition to normalizing data and potentially avoiding over-fitting, PCA also uncorrelates the variables, sometimes improving the performance of other modeling techniques.

Let's see if PCA improves or degrades the performance of hierarchical clustering.

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward's criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.


```{r}

# PC1 - PC7 capture 90% of the variance
wisc.pr.hclust <- hclust( dist(wisc.pr$x[,1:7]), method ="ward.D2" )
plot(wisc.pr.hclust)

```

```{r}

grps <- cutree(wisc.pr.hclust, k=2)
table(grps)

```

```{r}

table(grps, diagnosis)

```

```{r}

plot(wisc.pr$x[,1:2], col=grps)

```

```{r}

plot(wisc.pr$x[,1:2], col=diagnosis)

```

Note the color swap here as the hclust cluster 1 is mostly "M" and cluster 2 is mostly "B" as we saw from the results of calling table(grps, diagnosis). To match things up we can turn our groups into a factor and reorder the levels so cluster 2 comes first

```{r}

g <- as.factor(grps)
levels(g)

```

```{r}

g <- relevel(g,2)
levels(g)

```

```{r}

# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)

```

OPTIONAL: Lets be fancy and look in 3D with the rgl package we learned about in a previous class. Feel free to skip this optional step if you have difficulty installing the rgl package on windows machines. On mac you will likely need to install the XQuartz package from here: https://www.xquartz.org

```{r}

#library(rgl)
#plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
#rglwidget(width = 400, height = 400)

```

## Prediction

We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}

#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc

```

```{r}

# The two new patients are plotted onto the grapgh
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")

```

Patient 2 falls into the "malignant" group in the plot, so they should be prioritized for treatment.




