---
title: 'Class 06: R Functions'
author: "Madeline R. Luth"
date: "4/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview
Today we will focus on **R** functions but we will start with some **file reading**.

```{r}
plot(1:10, type = "l", col = "blue")
```
```{r}
read.table("test1.txt", header = TRUE, sep = ",")
```

Read in first file which is delimited by ","
```{r}
read.table("test2.txt", header = TRUE, sep = "$")
```
Read in second file which is delimited by "$"
```{r}
read.table("test3.txt")
```
Read in final file which has spaces and tabs

Note: Option + Command + i is shortcut to insert a new R chunk
```{r}
read.csv("https://bioboot.github.io/bggn213_W19/class-material/test2.txt")
```
Can also read.csv from a web domain instead of downloading file to the working directory

##  Our first function

Add some numbers
```{r}
add <- function(x, y=1) {
  #The body of the function
  x + y
}
```
```{r}
add(4)
```
```{r}
add(4,5)
```
Adding the 5 for y will override the default argument of y=1
```{r}
add(c(1,3,5), 1)
```
```{r}
#add(1,3,5)
#returns error of unused argument
```
```{r}
#add(x=1, y="barry")
#error because second value is non-numeric
```
```{r}
rescale <- function(x) {
   rng <-range(x)
   (x - rng[1]) / (rng[2] - rng[1])
}
```
```{r}
rescale( 1:10 )
```
```{r}
rescale( c(1,3,NA, 5, 10) )
```
```{r}
rescale2 <- function(x, na.rm=TRUE) {
  rng <- range(x, na.rm=na.rm)
  (x - rng[1]) / (rng[2] - rng[1])
}
```
```{r}
rescale2( c(1,3,NA, 10))
```
```{r}
rescale3 <- function(x, na.rm=TRUE, plot=FALSE) {
   if(na.rm) {
     rng <-range(x, na.rm=na.rm)
   } else {
     rng <-range(x)
   }
   print("Hello")
   answer <- (x - rng[1]) / (rng[2] - rng[1])
   print("is it me you are looking for?")
   if(plot) {
      plot(answer, typ="b", lwd=4)
}
   print("I can see it in ...")
}
```
This is another example
```{r}
rescale3( 1:10 )
```
```{r}
rescale3( 1:10, plot = TRUE)
```

